\section{Methods}

\subsection*{The rule}

Every cell carries a real-valued state vector $s \in \mathbb{R}^{C}$; the same parameters are shared by all cells and all developmental steps. Channel~0 is the signal/readout channel (clamped at input ports, read at output ports); the remaining channels are hidden. At each step a cell perceives four fixed features per channel from its four orthogonal neighbours---identity, the two central-difference gradients, and the discrete Laplacian:
\begin{equation}
\mathrm{perceive}(c) = \Big[\, s_c,\ \tfrac{1}{2}(s^{\to}_c - s^{\leftarrow}_c),\ \tfrac{1}{2}(s^{\downarrow}_c - s^{\uparrow}_c),\ s^{\to}_c + s^{\leftarrow}_c + s^{\uparrow}_c + s^{\downarrow}_c - 4 s_c \,\Big],
\end{equation}
concatenated over the $C$ channels into a vector of length $\mathrm{PERC}=4C$. These filters are fixed, not learned; they give the rule a sense of direction and curvature without any coordinate input, following the perception stencil of Growing Neural Cellular Automata \citep{mordvintsev2020growing}. The perceived vector drives a small two-layer perceptron with a single hidden ReLU layer of width $\mathrm{HD}$, and the output is added residually \emph{inside} a $\tanh$:
\begin{align}
\Delta &= W_2\,\mathrm{ReLU}\!\left(W_1\,\mathrm{perceive} + b_1\right) + b_2, \\
s' &= \tanh\!\left(s + \Delta\right).
\end{align}
Placing the residual inside the $\tanh$ (rather than the more common additive $s' = s + \Delta$) is a per-step nonlinearity that keeps every state bounded in $(-1,1)$ and lets the dynamics form fixed-point attractors. Input bits are re-clamped into the signal channel after every step, so the input acts as a persistent boundary condition; border cells are held at zero; damage zeros a cell's entire state, after which the input clamp re-applies. A single reference implementation (\texttt{rule.ts}) is the source of truth and is reproduced bit-for-bit by the GPU kernel and the fixed-point/Z80 datapath (below).

\subsection*{Configurations and initialisation}

One code path (\texttt{RuleConfig}) instantiates every experiment by choosing the grid size, channel count $C$, and hidden width $\mathrm{HD}$; the parameter block is laid out as $[\,W_1\,(\mathrm{HD}\times 4C),\ b_1\,(\mathrm{HD}),\ W_2\,(C\times \mathrm{HD}),\ b_2\,(C)\,]$. The XOR gate and the self-repair/grow-from-seed rules use a $9\times9$ grid with $C=12$, $\mathrm{HD}=48$ ($P=2940$ parameters); the 1-bit adder uses $11\times11$, $C=16$, $\mathrm{HD}=64$ ($P=5200$); the movable wire uses $17\times17$, $C=16$, $\mathrm{HD}=96$ ($P=7792$); the movable XOR uses $17\times17$, $C=20$, $\mathrm{HD}=128$ ($P\approx1.3\times10^{4}$). Parameters are initialised i.i.d.\ uniform on $[-0.06, 0.06]$ with the output layer ($W_2, b_2$) scaled by an additional factor of $0.5$, so the fresh substrate has weak but nonzero dynamics from the first step. The developmental rollout runs for $T$ steps ($T=24$ for the gate, $30$ for the adder, $40$--$50$ for the movable rules), where $T$ is chosen so the readout has settled.

\subsection*{Training by backpropagation through development}

The rule is trained by backpropagation through time \citep{werbos1990bptt} applied to the developmental rollout: we unroll $T$ steps of the cellular automaton, read a mean-squared-error loss at the output cell(s) of the final state (optionally averaged over a hold window; see below), and backpropagate the loss through every step to the single shared parameter vector. The forward pass stores the full state trajectory $s^{(0)},\dots,s^{(T)}$; the reverse pass then walks it backwards, at each step propagating the adjoint through the $\tanh$ ($\partial \tanh(z)/\partial z = 1 - \tanh^2(z)$, evaluated at the stored post-activation), through $W_2$, through the ReLU mask (gated by the stored pre-activation sign), through $W_1$, and finally through the fixed perception filters back onto the four neighbours' adjoints---accumulating the parameter gradient over all cells and all steps. Because the input clamp and the markers overwrite specific channels each step, their adjoints are zeroed at the corresponding entries before the step's gradient is propagated, so no gradient flows through the clamp.

Optimisation uses Adam ($\beta_1=0.9$, $\beta_2=0.999$, $\epsilon=10^{-8}$, with bias correction) and a small $\ell_2$ weight decay of $2\times10^{-5}$. The global gradient norm is clipped to $1$. The learning rate follows a linear warm-up over the first 20 iterations multiplied by a cosine decay from a peak to a floor over the run; from-scratch stages use a peak/floor of $0.01/0.004$ and warm-started stages a gentler $0.003/0.0006$, since a warm start is refinement rather than exploration and a hot rate would kick a good solution out of its basin. We keep the best parameters seen: the loss is evaluated and the parameter snapshot recorded \emph{before} each update, and the lowest-loss snapshot is returned. Because the XOR objective has a strong constant-$0.5$ local minimum, from-scratch stages use several random restarts (six for the gate's first curriculum stage) and keep the best.

\subsection*{Curricula}

Four curricula bootstrap otherwise hard-to-reach optima. \emph{Distance ramp}: long-range transport is a vanishing-gradient problem from scratch, so the output cell starts adjacent to the inputs and is moved one cell farther each stage, each stage warm-started from the previous solution, until the target separation is reached. \emph{Persistence (hold window)}: rather than scoring a single readout step, the output MSE is averaged over the last $w$ states of the rollout, which makes the correct answer a temporal attractor the field must hold rather than pass through:
\begin{equation}
\mathcal{L} = \frac{1}{|\mathcal{B}|\,w}\sum_{\text{samples}}\ \sum_{t=T-w+1}^{T}\ \sum_{k}\big(s^{(t)}_{o_k,0} - y_k\big)^2 .
\end{equation}
\emph{Reactivity}: an input bit is flipped mid-rollout at a switch step, and a second hold window scores re-settling to the new answer, training the field to track live input changes without re-seeding. \emph{Expose-then-internalise}: to give the 2-bit ripple-carry adder a produced-then-consumed internal carry, the intermediate carry is first teacher-forced (exposed) and then released, so the field must both produce and consume it internally.

\subsection*{Position invariance via re-stamped markers}

Movable rules read \emph{ports}, not coordinates. Two dedicated marker channels encode input ports (channel~1) and output ports (channel~2). Every step, after the rule is applied everywhere, all marker channels are cleared and re-stamped at the (possibly relocated) port positions, and the input bits are re-injected there. Because the rule sees only marker activity and its local neighbourhood---never an absolute coordinate---the same parameters route and compute for arbitrary port placements and, empirically, for grid sizes never seen in training. Ports may be dragged at test time and the field re-routes under fixed parameters.

\subsection*{Separated signal channels}

For position-invariant computation with two inputs (movable XOR), the two input bits are injected into \emph{distinct} signal channels (channels 3 and 4) while channel~0 remains the readout, rather than a single shared channel. A shared channel superimposes the two travelling bit-waves so the output cannot decode which input was which and the objective collapses to the constant-$0.5$ baseline; separate channels keep the bits decodable and the task learnable. This is a representational constraint, not a capacity one.

\subsection*{Stochastic (asynchronous) updates}

Synchronous update admits a period-3 synchronous limit cycle (a pole near $\pm120^\circ$) that manifests as a blinking readout. We optionally desynchronise with NCA-style stochastic updates: each cell updates with probability $\mathrm{fireRate}<1$ per step and otherwise keeps its state, damping the limit cycle into a genuine fixed point (the reactive movable XOR uses $\mathrm{fireRate}=0.5$). The per-cell fire mask is a deterministic hash of (cell, step, seed) that is bit-identical between the CPU reference and the WGSL kernel, so the asynchronous rollout and its gradient validate exactly across implementations; in the backward pass a non-firing cell is the identity map, so its adjoint passes straight through with no parameter contribution.

\subsection*{GPU trainer and cross-validation}

For affordable multi-seed statistics we implemented an in-browser WebGPU trainer that performs the same reverse-mode BPTT in a batch-packed WGSL kernel. Its gradient matches the f64 reference to $\approx 3\times10^{-5}$ (max.\ relative error, movable/marker rules) and it runs $\approx 65\times$ faster than the CPU trainer (for example, training the 1-bit adder to 8/8 cases in $\approx37$~s). The CPU reference (\texttt{lossAndGradFixed} for fixed-layout multi-input/multi-output rules; \texttt{lossAndGradMarkers} for movable rules) is itself finite-difference-checked, and the GPU path is validated against it to a maximum relative error of $1.97\times10^{-6}$. All three implementations---the f64 reference, the WGSL kernel, and the fixed-point/Z80 datapath---are held to agree; the discipline is that the emulator is never validated against model code, only the reverse: the reference is checked by finite differences and the substrate by the reference. We note that headless WebGPU was unavailable in our environment, so GPU multi-seed runs were driven in-browser.

\subsection*{The Z80 fixed-point substrate}

To close the loop to a real instruction set, the trained rule is quantised to signed fixed-point and executed on a conformance-tested Z80 emulator. A bit-faithful fixed-point emulation of the rule (\texttt{fixed.ts}) fixes the datapath's precision, and that exact datapath is implemented in Z80 assembly and run on \texttt{z80-emulator} v2.3.0 (a TypeScript implementation of the Z80 ISA). This is an emulator of the instruction set, not physical silicon, and it is a different emulator from the \texttt{superzazu} core against which our separate GPU Z80 was conformance-tested. The layout is ``grid$\to$lane'': one lane holds the field and weights; because the target GPU core has no cross-lane communication, this is an \emph{offline provenance proof} rather than the live demo. Validation proceeds bottom-up: the signed Q8.8 multiply-accumulate is bit-exact against the reference over 300 random vectors and all 48 rows of $W_1$; one full cell update is bit-exact over 144 real cell updates ($\approx2\times10^{-2}$ versus f64); and the whole XOR gate---each cell update executed on the emulator, with the grid sweep and 24-step rollout orchestrated (and memoised) in TypeScript---reproduces the grown truth table $0.05/0.98/0.99/0.04$ at Q8.8. The assembler was extended with CB-prefixed \texttt{SRA}/\texttt{RR} and a strict \texttt{JR}/\texttt{DJNZ} range check; all changes are backward-compatible (the prior dual-number and CA differential-test suites pass with zero diff).

\subsection*{In-substrate forward-mode gradient}

The same integer datapath can additionally carry a gradient. We implement forward-mode automatic differentiation \citep{wengert1964ad,griewank2008evaluating} as dual numbers in fixed point: each primal value is paired with a tangent, and the tangent is propagated alongside the value through the multiply-accumulate, the ReLU, and the $\tanh$ of a single cell update. On the Z80 this reproduces the derivative $\mathrm{d}(\text{output})/\mathrm{d}\theta$ and matches finite differences to $4.3\times10^{-3}$ at Q8.8; in TypeScript the same fixed-point tangent is clean at Q16.16 ($1.5\times10^{-5}$). We state its scope precisely: this is a single-cell, single-step Jacobian-vector product (JVP) for a handful of weights and one input case---the differentiation \emph{primitive} running co-resident with the value computation in one integer instruction set---\emph{not} the full training gradient $\partial\mathcal{L}/\partial\theta$, which is the reverse-mode BPTT gradient over $T$ steps, all interior cells, and all input cases described above. The in-substrate result demonstrates the mechanism (a learned rule and a matching tangent in one real ISA); recovering the training gradient in-substrate would require propagating the forward tangent through the full rollout, or equivalently the reverse-mode adjoint, rebuilt at Q16.16.

\subsection*{Statistics}

Ablations and reproducibility runs follow a fixed protocol: 8 random seeds per condition (16 for load-bearing claims), with success defined as every truth-table case landing within a fixed tolerance of its target ($0.2$ for the gate, $0.3$ for the adder and movable tasks; we report the threshold and its sensitivity). Success rates are reported with Wilson 95\% confidence intervals and continuous metrics as mean\,$\pm$\,s.d. Under this protocol the XOR gate solves 8/8 seeds (mean loss $1.0\times10^{-5}\pm8.0\times10^{-6}$); an identity-only perception control solves 0/8 (loss exactly $0.25$, the constant-output baseline), confirming that spatial coupling is necessary; removing the hidden ReLU solves 2/8; and a hidden-width sweep saturates early (5/8 at $\mathrm{HD}=4$, 8/8 from $\mathrm{HD}=8$ upward, though 5/8-versus-8/8 is not statistically separated at $n=8$). The 1-bit adder reproduces at 6/8 seeds (Wilson 95\% $[41\%, 93\%]$) and the movable wire at 9/10 seeds ($\geq90\%$ routing accuracy, mean $98\pm4\%$; Wilson 95\% $[60\%, 98\%]$). We are explicit that, apart from these, most headline results---self-repair, grow-from-seed, the reactive 1-bit adder, the 2-bit adder, and the movable XOR---are at present single trained models ($n=1$) and should be read as existence proofs rather than measured success rates; position-invariant \emph{computation} in particular reaches only $\approx68\%$ at $17\times17$.
