\section{Results}

\subsection{A learned local rule grows a functional gate}

We study a single update rule applied identically at every cell of a two-dimensional grid. Each cell holds a vector of real-valued state channels. At every developmental step the rule \emph{perceives}, for each channel, four fixed features of the cell's four-neighbour stencil---the cell's own value (identity), the two first spatial derivatives $\partial_x$ and $\partial_y$, and the discrete Laplacian $\nabla^2$---concatenates them, and maps them through a small multilayer perceptron (MLP) with one hidden rectified-linear (ReLU) layer. The MLP output is added to the current state and squashed:
\begin{equation}
  \mathbf{s}' = \tanh\!\big(\mathbf{s} + W_2\,\mathrm{relu}(W_1\,\mathrm{perceive}(\mathbf{s}) + \mathbf{b}_1) + \mathbf{b}_2\big).
  \label{eq:rule}
\end{equation}
The residual sum sits \emph{inside} the $\tanh$, bounding every channel to $(-1,1)$ and allowing the dynamics to settle onto attractors rather than diverge. The perception filters are fixed rather than learned; only the MLP weights $(W_1,\mathbf{b}_1,W_2,\mathbf{b}_2)$ are trained, and they are shared across every cell and every step (\ref{fig:method}a,b). Input values are re-clamped into a designated signal channel at each step and border cells are held at zero.

We train this rule by backpropagation through time (BPTT) \citep{werbos1990bptt} over a rollout of $T$ developmental steps, reading a loss only at designated output cells at the end of the rollout. This is the same rule form and developmental-training principle introduced for growing images \citep{mordvintsev2020growing}, but the target is a truth table read at an output cell rather than a fixed picture.

The first target is a two-input XOR gate on a $9\times9$ grid, with the output cell placed five cells away from the inputs and moved progressively outward over training (a distance curriculum). The trained rule grows a field that routes both input waves to the output cell and computes their exclusive-or: across the four input cases the output cell reads $0.05$, $0.98$, $0.99$ and $0.04$, matching the target truth table (\ref{fig:hero}b, \ref{fig:compose}a). This is reproducible rather than a single fortunate run: across eight independently seeded trainings, all eight solve the task ($8/8$; success defined as all four cases within $0.2$ of target), at a mean final loss of $1.0\times10^{-5} \pm 8.0\times10^{-6}$.

Ablations isolate what the rule requires. Restricting perception to the identity feature---removing all neighbour information---collapses success to $0/8$ at a loss of exactly $0.2500$, the mean-squared error of emitting a constant output; without spatial coupling the displaced output cell is structurally severed from the inputs. Removing the hidden ReLU (a linear hidden layer, with the per-step $\tanh$ retained) drops success to $2/8$. Hidden width saturates early: $5/8$ solve at width $4$ versus $8/8$ at width $8$ and above, although at $n=8$ the width-$4$ versus width-$8$ gap is not statistically separated. Directional perception, by contrast, is redundant on this symmetric gate---an isotropic identity-plus-Laplacian rule also solves $8/8$---a point we return to once port placement is free (\ref{fig:invariance}).

\subsection{The paradigm composes into arithmetic}

The gate is the base of a ladder. Holding the rule form of Eq.~\eqref{eq:rule} fixed and training fresh weights for each task---the capabilities reported here are, at present, separate trained rules rather than one universal rule---we grow arithmetic circuits of increasing compositional depth (\ref{fig:compose}).

A one-bit full adder takes three inputs (two addends and a carry-in) and produces a sum and a carry-out at two separate output cells. A trained rule computes all eight rows of its truth table. The result reproduces across seeds, though less tightly than the gate: six of eight independently seeded trainings reach the success bar ($6/8$; Wilson $95\%$ confidence interval $[41\%, 93\%]$). The same rule tracks live inputs, re-settling to the new sum and carry across all $64$ input transitions when an input is flipped mid-rollout---a property we develop in \ref{fig:robust}.

The two-bit ripple-carry adder is the paradigm's first test of genuine compositional depth, because the carry produced by the low-order bit must be consumed by the high-order bit \emph{within} the field, with no external wiring. A single rule solves all sixteen cases ($16/16$). We established that this carry is genuinely internal, and not smuggled through the outputs, in two ways. First, a curriculum that teacher-forces an internal carry signal and then releases it forces the field to both produce and consume the carry on its own. Second, a causal lesion: zeroing the internal carry region leaves the carry-\emph{independent} low bit intact (sum$_0$: $16/16$) while degrading exactly the carry-\emph{dependent} outputs (sum$_1$ and carry-out: $10/16$)---a clean functional dissociation that localises the carry's causal role (\ref{fig:compose}c). Two caveats are stated plainly. The carry is represented in a distributed way, so single-cell interventions are only partial and the dissociation requires lesioning the whole carry region; and the two-bit adder is at present a single trained seed ($n=1$), the largest computation we grow. Unlike a monolithic differentiable arithmetic learner \citep{kaiser2016neuralgpu}, the circuit here is assembled developmentally from a local rule and carries its intermediate signal in the substrate itself.

\subsection{Robust to damage and asynchronous update}

A grown computer is only useful if it survives insult. Training the rule with a damage stage---zeroing a square patch of cells mid-rollout and continuing to score the output---makes self-repair an earned property of the dynamics. After a patch of the settled field is destroyed, the rule regrows the missing structure in roughly twenty steps and the output returns to the correct answer and then holds it (\ref{fig:hero}b, \ref{fig:robust}a). The same rule form grows the entire circuit from a single seed cell and settles onto an attractor that persists beyond $600$ developmental steps. This regeneration toward a stored functional target is the engineered analogue of biological pattern memory in regeneration \citep{durant2017longterm}, and inherits the self-repair lineage of growing neural cellular automata \citep{mordvintsev2020growing}. As above, the grow-from-seed and self-repairing rules are separately trained instances ($n=1$).

Because the rule is applied at every cell, the order of updates matters. Under fully synchronous update the settled field is not a fixed point but a period-three limit cycle---every cell blinks with period three---which a spectral analysis of the state autocorrelation makes explicit. Stochastic asynchronous update, in which each cell applies the rule with probability below one on each tick (fire rate $<1$, as in asynchronous neural cellular automata \citep{mordvintsev2020growing}), damps this limit cycle into a genuine fixed point; a sweep over the fire rate traces the transition (\ref{fig:robust}b). The asynchronous fire mask is bit-identical across our CPU and GPU implementations, so the effect is a property of the dynamics rather than of a particular scheduler.

The settled computation is stable over long horizons: output accuracy is unchanged when the rollout is extended to $50$, $150$ and $400$ steps, and the field remains an attractor well beyond (\ref{fig:robust}c). Finally, the circuit is reactive rather than a one-shot developmental artefact: flipping an input of the one-bit adder mid-rollout, with no re-seeding, the output cells re-settle to the new sum and carry across all $64$ input transitions (\ref{fig:robust}d).

\subsection{One rule computes at any placement and grid size}
\label{sec:invariance}

The rule reads re-stamped marker channels rather than absolute coordinates (Methods), so the input and output ports can be moved to arbitrary locations and the grid resized without retraining. This raises a question that separates routing from computation: does a single set of learned parameters transport and combine information correctly when the geometry it was never trained on changes?

For \emph{routing}---moving one bit from an input port to a displaced output port---the answer is an unqualified yes. A single trained parameter set routes correctly at $100\%$ accuracy on $11\times11$, $13\times13$, and $17\times17$ grids under identical parameters (Fig.~\ref{fig:invariance}b). Trained from scratch over multiple seeds, the movable-wire rule reproduces: $9$ of $10$ seeds reach $\ge 90\%$ routing accuracy over randomly placed ports at $17\times17$ (mean $98\pm4\%$; Wilson $95\%$ CI $[60\%,98\%]$; $n=10$; Fig.~\ref{fig:invariance}d). Waves emitted from the ports meet and lock into a stable conduction path regardless of where the ports are stamped.

For \emph{computation}---a movable XOR, where two input bits must be routed \emph{and} combined at a displaced output---placement invariance is only partial. A single trained rule reaches $100\%$ at $11\times11$, $95\%$ at $13\times13$, but degrades to $68\%$ at $17\times17$ (single seed); a reactive variant that must re-settle after a live input flip sits near the chance floor at $\sim\!58\%$ and is effectively unsolved. We report these numbers plainly: position-invariant multi-input computation at scale is the open crux of this work, not a solved capability.

The gap between routing and computation exposes a representational rather than a capacity constraint. Movable XOR is learnable only when each input bit occupies its own signal channel; with a shared channel the two bit-waves superpose and the loss plateaus exactly at the constant-0.5-output baseline (MSE $0.25$), no matter the hidden width (Fig.~\ref{fig:invariance}c). Separating the channels---not adding capacity---is what makes the placed computation trainable. This channel-separation boundary is, to our reading, the more general finding: local developmental rules generalise routing across placement and scale for free, but combining several placed inputs requires that the substrate keep those inputs representationally distinct.

The invariance demonstrated here is spatial: the same rule computes under translation of its I/O and rescaling of its grid. This is a different axis from the input-length or problem-size generalisation obtained by running a recurrent computation for more iterations \citep{schwarzschild2021easytohard} or by architectures that length-generalise arithmetic \citep{kaiser2016neuralgpu}, neither of which grows, is placed, or self-repairs.

\subsection{The learned rule is a program in a real instruction set}
\label{sec:substrate}

A learned local rule need not remain a floating-point abstraction. Quantised to signed fixed-point integers, the trained gate rule becomes an ordinary program that executes on a real Z80 instruction-set architecture. We fix the datapath in a bit-faithful fixed-point emulation of the rule (perceive $\rightarrow$ signed multiply--accumulate $\rightarrow$ tanh lookup) and implement that exact datapath in Z80 assembly, run on a third-party software emulator of the ISA (\texttt{z80-emulator} v2.3.0). This is an offline provenance proof---each cell update is executed on the emulated Z80, while the grid sweep and the $24$-step developmental rollout are orchestrated in software---and it runs on an emulator, not physical silicon. Within that scope the result is exact: across $144$ real cell updates the Z80 output is bit-identical to the fixed-point reference (and within $\approx\!2\times10^{-2}$ of the $f64$ rule), and the grown gate computes the correct truth table entirely in the substrate, reading $0.05/0.98/0.99/0.04$ at Q8.8 (Fig.~\ref{fig:substrate}b). The same datapath is cross-validated bit-for-bit against the $f64$ reference and the WGSL GPU kernel.

The same integer datapath also carries a derivative. Alongside the value computation we propagate a forward-mode (dual-number) tangent \citep{wengert1964ad,griewank2008evaluating} through the fixed-point arithmetic, and it matches finite differences on the Z80 to $4.3\times10^{-3}$ at Q8.8 and to $1.5\times10^{-5}$ at Q16.16 (Fig.~\ref{fig:substrate}c). We state precisely what this tangent is and is not. It is a single-cell, single-step Jacobian--vector product---the forward-mode sensitivity of one cell's update to six hand-picked weights for one input case. It is \emph{not} the reverse-mode training gradient $\partial L/\partial\theta$, which is computed by backpropagation through the full $T$-step rollout over all cells and cases \citep{werbos1990bptt} and is what actually trains the rule. The contribution is the co-residence: a learned rule and a matching forward-mode gradient realised in the same real integer datapath (Fig.~\ref{fig:substrate}a,d). This bridge distinguishes the result from physical and in-materia learning substrates, which train an analog device as a fixed inference artifact \citep{wright2022deep,prezioso2015training}; here the learned rule is itself an exact, reproducible program that carries its own tangent.

\subsection{Ablations, baselines, and statistics}
\label{sec:ablation}

We hold the gate to a multi-seed standard and use it as the template for the rest of the work. On the $9\times9$ XOR gate with a distance curriculum to $d=5$ (success $=$ all four cases within $0.2$ of target), the full rule solves $8/8$ seeds at a mean loss of $1.0\times10^{-5}\pm8.0\times10^{-6}$ (Fig.~\ref{fig:ablation}a). Three ablations isolate what the rule needs.

\emph{Spatial coupling is necessary.} Replacing the perception stencil with identity only---each cell sees itself but not its neighbours---collapses success to $0/8$ at a mean loss of exactly $0.25$ ($\pm6\times10^{-17}$). This is precisely the mean-squared error of a constant output: without neighbour information the displaced output cell is structurally disconnected from the inputs and can do no better than ignoring them. The control confirms the harness and the necessity of coupling; it is partly tautological and we present it as such.

\emph{The hidden nonlinearity matters.} Removing the hidden ReLU (a linear hidden layer, though the per-step tanh remains) drops success to $2/8$ at a mean loss of $0.092\pm0.10$: one of the two nonlinearities is load-bearing, but the tanh alone still solves the task for some seeds.

\emph{Capacity saturates early.} A hidden-width sweep gives $5/8$ at HD$=4$ and $8/8$ at HD$=8,16,32,96$, placing the threshold near $8$. We caution that $5/8$ versus $8/8$ is not statistically separated (Fisher $p\approx0.2$), so ``threshold $\approx 8$'' is a point estimate at $n=8$, not a measured boundary. Directional perception is redundant on this symmetric gate (isotropic perception also solves $8/8$); the ablation that should expose directionality belongs on the movable task and is pending.

Multi-seed statistics currently reach three results: the gate ($8/8$, $n=8$), the 1-bit full adder ($6/8$ from scratch; Wilson $95\%$ CI $[41\%,93\%]$), and the movable wire ($9/10$; \S\ref{sec:invariance}). Every other headline in this paper---the 2-bit ripple-carry adder, movable XOR, self-repair, and grow-from-seed---is a single trained model and should be read as an existence proof rather than a measured success rate, a limitation we state in the main text. Two ablations that would make the story airtight are also outstanding: the directional-perception ablation on the movable task (where directionality should be strictly necessary), and a non-developmental baseline---a global MLP trained on the same truth tables---to isolate exactly what development buys along the capability axes (self-repair, position and scale invariance, ISA execution) where only the developmental rule is expected to succeed \citep{mordvintsev2020growing}. We report the completed ablations as the rigor floor and name the missing controls rather than paper over them.
